Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-09-08 docs/papers/shallow_performance/shallow_perf.tex
parco paper: correct my psykal macro to not insert trailing blank space

\newcommand{\psykal}{{PS}y{KA}l\ }

@@

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see significant
speed-ups) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system.
%%
Therefore, the \psykal approach can be used with negligable
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the PSyKAl version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

\subsection*{The \psykal Approach}

The \psykal approach attempts to address the problems described in the
previous section. It separates code into three layers; the Algorithm
layer, the PSy layer and the Kernel layer. The approach has been
developed in the GungHo project~\cite{GungHo}, which is creating a new
Dynamical core for the Met Office, and its design has been influenced
by earlier work on OP2~\cite{OP2,PYOP2}.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of Shallow retains the Input/Output
capability of the original (it reads the problem specification from a
namelist and outputs the solution fields every $n$ timesteps). This
aids in limiting the optimisations that can be performed on the
\psykal version to those that should also be applicable to full
oceanographic models. Note that although we retain the I/O
functionality, all of the results presented in this work carefully
exclude the effects of I/O since it is compute performance that
interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal version of Shallow, we made the kernels
as fine-grained as possible. In this case, this resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels). With a little
bit of tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
routine then consists of applying each of the kernels to all of the
points on the model mesh requiring an update. In its basic,
unoptimised (`vanilla') form, this PSy-layer routine then contains a
doubly-nested loop around each kernel call, as indicated by the
pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.

\begin{figure}
\centering
\includegraphics[width=100mm]{psykal_shallow}
\caption{A schematic of the vanilla \psykal version of the Shallow code.}
\label{FIG_psykal_shallow_structure}
\end{figure}

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. In the case of Shallow we
have periodic boundary conditions in both the $x$ and $y$ dimensions.
These are simply implemented by having additional (`halo')
rows/columns at the edges of the domain and ensuring the data in them
is up-to-date before it is read. These rows/columns are updated by
copying data from the corresponding row/column on the opposite side of
the domain. In the current work, the lines of code to do these copies
are manually inserted in the PSy-layer routine as required.

Excluding the application of boundary conditions, the code has three
main sections: computation of mass fluxes, vorticity and sea-surface
height; computation of the pressure and velocity components; and
finally, smoothing of the fields in time and copying of the field
values at the next time-step to be those at the current
time-step. Profiling of the original code showed that these three
sections each account for approximately one third of the run-time.

@@

The primary aim of this work is to determine whether any performance
loss due to the \psykal separation of concerns can be negated by
optimising the PSy layer and if so, what the key transformations are
(for a given system). To evaluate this we perform tests on a range of
hardware and a range of compilers. The hardware and compilers are
listed in Table~\ref{TABLE_compilers}. Where a compiler is available
on a given piece of hardware, the version number used in this work is
specified.
%
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
`chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
3.3~GHz.

@@

In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
used with each compiler. These flags are particularly important for
the \psykal versions of the code. The performance of the original
version of the code is much less sensitive to the compiler
options. This is because it consists of a single source file
containing effectively a single routine (excluding those for I/O).
For all compilers apart from Cray it was important to use flags that
encourage in-lining. With the Cray compiler some benefit was obtained
by increasing the level of Inter-Procedural Analysis and using
'whole-program optimisation'. In order to get performant code from the
IBM compiler it was found essential to make the pre-fetching
considerably more aggressive than the default.

@@

Beginning with the vanilla \psykal version, we then manually apply a
series of (cumulative) code transformations while obeying the \psykal
separation of concerns, {\it i.e.} optimisation is restricted to the
middle, {PS}y layer and leaves the kernel and algorithm layers
unchanged. The aim of these optimisations is to recover, as much as is
possible, the structure and thus performance of the original version
of the code. We began with two transformations that affect all of the
kernels. After this, we determined the transformations to perform by
comparing the timing profiles of the \psykal and original versions of
the code. We then tackled the sections resposible for the largest
slowdowns and iterated this process. While doing this we also made
heavy use of the optimisation reports produced by the compilers and
how these changed when moving from the original to the \psykal version
of the code.

@@

\item Move all kernel subroutines into the same module as as that
  containing the middle/PSy layer (`module inlining');

@@

We explore the extent to which performance depends upon the problem
size by using square domains of dimension 64, 128, 256, 512 and
1024. This range allows us to investigate what happens when cache is
exhausted as well as giving us some insight into the decisions that
different compilers make when optimising the code.

@@

In this section we first discuss the performance of the original
version of the code with the various compiler/CPU combinations. We
then investigate the different code transformations that are required
to recover good performance from the \psykal version of Shallow. Finally,
we compare the most performant \psykal version  with the original
version of Shallow for each compiler/CPU combination.

@@

For this compiler-friendly form of the code, there is generally little
performance difference between the executables produced by the Gnu and
Intel compilers on both the Haswell and Ivy Bridge CPUs. However, the
executable produced by the Cray compiler generally performs
significantly better for all but the largest domain size. Analysis of
the optimisation reports produced by the Cray and Intel compilers did
not give any clues as to the cause of this difference and an in-depth
investigation is outside the scope of this paper.

\subsection{Optimisations of the \psykal Version to Recover Performance}

We now examine the performance of the \psykal version of Shallow and
the code transformations that are required to reach the levels obtained
by the original.  We do this for the $256^{2}$ problem size since this
fits within last-level cache on all of the CPUs we are using here.
The performance of the \psykal version after each transformation has
been applied to the code is plotted in
Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
given relative to the performance of the original version of Shallow
with the same compiler/CPU.

\begin{figure}[!t]
\centering
\includegraphics[width=123mm]{opt_stages_256}
\caption{The performance of the \psykal version of Shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called (module in-lining) has the
most dramatic effect on the performance of the compiled code; it now
achieves more than 80\% of the performance of the original. The
transformations required to get beyond this level then become
architecture dependent: on Haswell loop fusion achieves more than 90\%
while on Ivy Bridge it is fully-fusing the first loop nest. The
dramatic performance improvement obtained from the module-inlining
transformation demonstrates that the Gnu compiler is unable to
optimise over separate files.

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal version of
Shallow achieves some 80\% of the performance of the original
version. Only two code transformations were required to increase this
to $\sim$100\%. The first of these was the fusion of the computational
loops. However, we found that performance suffered (bar 5) when this
was extended to include the loop required for field copies at the end
of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
investigation of this slow-down revealed that the compiler was unable
to SIMD vectorise the newly-fused loop after it had in-lined the body
of the kernel called from within it. Once this in-lining was done by
hand (bar 6), the compiler was able to vectorise the loop and the
performance of the original version of Shallow was recovered on the
Haswell system. On Ivy Bridge the first loop nest also had to be fully
fused (bar 7) to recover $\sim100\%$ of the performance.

@@

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal version of Shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal version is fairly
good at 74\% of that of the original. In contrast to the other
compilers consisdered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal version that achieves 88\% of the
performance of the original.

Subsequent transformations actually harm performance until the
field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
with the preceeding loop. The resulting \psykal version now achieves
94\% of the original performance. Two further steps are required to
match the latter. First, the first loop nest has to be fully fused
(both inner and outer loops fused) and second, the array/loop bounds
are passed as an argument to the middle layer. This is done in order
to indicate to the compiler that these bounds remain constant for the
duration of each trip of the time-stepping loop.

Moving to the IBM compiler on the Power 8 system (sixth cluster in
Figure~\ref{FIG_opt_stages_256}) we see that performance of the
vanilla \psykal version is only 5\% of the original and several
transformations are required to achieve comparable performance. As
with the Cray compiler, specifying array bounds explicitly improves
performance. Module-inlining also has a significant effect; doubling
the performance from 10 to 20\% of the original. However, fusing loop
nests actually harms performance, just as it did for the Cray
compiler. Kernel inlining takes performance to over 50\% and it is the
IBM-supplied transformations of loop unrolling and fissioning the
first loop ({\it i.e.} undoing some of transformations three and
seven) that finally recovers 96\% of the performance of the
original. We note that this final pair of transformations significantly
reduce the performance obtained with the Intel and Gnu compilers
although they do not much perturb that obtained by the Cray
compiler.

\subsection{Performance Comparison of the Original and \psykal Versions}

Finally, we compare the performance of the best \psykal version of
Shallow with that of the original. In
Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
  difference} between the performance of the original and the (best)
\psykal versions of Shallow for each compiler/CPU combination. The
most important feature of this plot is that the performance penalty
incurred by the \psykal version is less than four percent with only
one exception where it is 6.6\%.  In fact, for the Intel and Cray
compilers, the \psykal version of Shallow is never more than 2\%
slower than the original and in some cases is faster.  This
demonstrates that we can reliably recover the performance of the
original version of the code, despite the significant restructuring
required by the \psykal approach.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=120mm]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal
version with that of the original version of the code. A negative value 
indicates that the \psykal version is faster than the original.}
\label{FIG_slowdown_summary}
\end{figure}

Turning to the cases where the \psykal version is significantly
{\em faster} than the original, we see from
Figure~\ref{FIG_slowdown_summary} that, with the exception of Power 8,
this only occurs for the smallest domain size. This is no doubt due in
part to the fact that the problem executes very quickly and therefore
a small reduction in run-time will show as a large precentage. Since
this problem is well contained in L2 cache, its performance will also
be more sensitive to small differences, {\it e.g.} in the way that CPU
registers are used.

The performance of the Power 8 system with the $512^2$ domain size is
somewhat anomolous: the \psykal version of the code performs
significantly better than the original and in addition, this problem
size is also the only one where the Power 8 out-performs the Haswell
system (see Figure~\ref{FIG_orig_perf_summary}) for the original
version.  The results in Figure~\ref{FIG_orig_perf_summary} indicate
that the $512^2$ case marks a transition from the working set being in
cache (sizes $\leq 256^2$) to spilling out of cache (size $1024^2$). We
therefore suggest that cache is being used more efficiently on the
Power 8 system and this is further enhanced by the IBM-supplied
transformations in the \psykal version.

@@

We have investigated the application of the \psykal separation of
concerns approach to the domain of finite-difference ocean
models. This approach enables the computational science (performance)
related aspects of a computer model to be kept separate from the
natural (oceanographic) science aspects.

@@

In particular, the results on the IBM system demonstrate two key points.
%
First, even for a relatively simple
benchmark code, the code structure required to obtain good performance
between different architecture and/or compiler combinations may
differ. For hand optimised codes this implies the need to support
multiple code bases in order to achieve portable performance.
%
Second, the \psykal approach directly addresses the problem of
requiring different optimisations for different architectures and/or
compilers by limiting these differences to a separate performance
layer, thereby allowing the natural science code base to remain
unchanged.

In future work we will extend our performance portability analysis of
the \psykal approach to shared memory parallelisation optimisations on
different architectures using different parallelisation techniques (in
particular OpenMP on multi/many-core and OpenACC on GPU's).
%
We will then analyse the performance of a domain-specific compiler
that is currently under development. This compiler will generate
optimised PSy-layer code using a user-provided recipe of
transformations, thereby removing the need for optimisation experts to
manually write the PSy layer.

>>>>

\newcommand{\psykal}{{PS}y{KA}l}

@@

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see significant
speed-ups) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system.
%%
Therefore, the \psykal\ approach can be used with negligable
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the PSyKAl version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

\subsection*{The \psykal\ Approach}

The \psykal\ approach attempts to address the problems described in the
previous section. It separates code into three layers; the Algorithm
layer, the PSy layer and the Kernel layer. The approach has been
developed in the GungHo project~\cite{GungHo}, which is creating a new
Dynamical core for the Met Office, and its design has been influenced
by earlier work on OP2~\cite{OP2,PYOP2}.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal\ version of Shallow retains the Input/Output
capability of the original (it reads the problem specification from a
namelist and outputs the solution fields every $n$ timesteps). This
aids in limiting the optimisations that can be performed on the
\psykal\ version to those that should also be applicable to full
oceanographic models. Note that although we retain the I/O
functionality, all of the results presented in this work carefully
exclude the effects of I/O since it is compute performance that
interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of Shallow, we made the kernels
as fine-grained as possible. In this case, this resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels). With a little
bit of tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
routine then consists of applying each of the kernels to all of the
points on the model mesh requiring an update. In its basic,
unoptimised (`vanilla') form, this PSy-layer routine then contains a
doubly-nested loop around each kernel call, as indicated by the
pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.

\begin{figure}
\centering
\includegraphics[width=100mm]{psykal_shallow}
\caption{A schematic of the vanilla \psykal\ version of the Shallow code.}
\label{FIG_psykal_shallow_structure}
\end{figure}

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. In the case of Shallow we
have periodic boundary conditions in both the $x$- and $y$-directions.
These are simply implemented by having additional (`halo')
rows/columns at the edges of the domain and ensuring the data in them
is up-to-date before it is read. These rows/columns are updated by
copying data from the corresponding row/column on the opposite side of
the domain. In the current work, the lines of code to do these copies
are manually inserted in the PSy-layer routine as required.

Excluding the application of boundary conditions, the code has three
main sections: computation of mass fluxes, vorticity and sea-surface
height; computation of the pressure and velocity components; and
finally, smoothing of the fields in time and copying of the field
values at the next time-step to be those at the current
time-step. Profiling of the original code showed that these 
sections each account for approximately one third of the run-time.

@@

The primary aim of this work is to determine whether any performance
loss due to the \psykal\ separation of concerns can be negated by
optimising the PSy layer and if so, what the key transformations are
(for a given system). To evaluate this we perform tests on a range of
hardware and a range of compilers. The hardware and compilers are
listed in Table~\ref{TABLE_compilers}. Where a compiler is available
on a given piece of hardware, the version number used in this work is
specified.
%
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
`chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
3.3~GHz.

@@

In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
used with each compiler. These flags are particularly important for
the \psykal\ versions of the code. The performance of the original
version of the code is much less sensitive to the compiler
options. This is because it consists of a single source file
containing effectively a single routine (excluding those for I/O).
For all compilers apart from Cray it was important to use flags that
encourage in-lining. With the Cray compiler some benefit was obtained
by increasing the level of Inter-Procedural Analysis and using
'whole-program optimisation'. In order to get performant code from the
IBM compiler it was found essential to make the pre-fetching
considerably more aggressive than the default.

@@

Beginning with the vanilla \psykal\ version, we then manually apply a
series of (cumulative) code transformations while obeying the \psykal
separation of concerns, {\it i.e.} optimisation is restricted to the
middle, {PS}y layer and leaves the kernel and algorithm layers
unchanged. The aim of these optimisations is to recover, as much as is
possible, the structure and thus performance of the original version
of the code. We began with two transformations that affect all of the
kernels. After this, we determined the transformations to perform by
comparing the timing profiles of the \psykal\ and original versions of
the code. We then tackled the sections resposible for the largest
slowdowns and iterated this process. While doing this we also made
heavy use of the optimisation reports produced by the compilers and
how these changed when moving from the original to the \psykal\ version
of the code.

@@

\item Move all kernel subroutines into the same module as that
  containing the middle/PSy layer (`module inlining');

@@

We explore the extent to which performance depends upon the problem
size by using square domains with side lengths of 64, 128, 256, 512 and
1024. This range allows us to investigate what happens when cache is
exhausted as well as giving us some insight into the decisions that
different compilers make when optimising the code.

@@

In this section we first discuss the performance of the original
version of the code with the various compiler/CPU combinations. We
then investigate the different code transformations that are required
to recover good performance from the \psykal\ version of Shallow. Finally,
we compare the most performant \psykal\ version  with the original
version of Shallow for each compiler/CPU combination.

@@

Strangely, the third bar (Gnu compiler) in
Figure~\ref{FIG_orig_perf_summary} shows the same drop in performance
in moving from the $256^2$ case to the $512^2$ case, despite the
larger last-level cache of the Ivy Bridge system. In order to
investigate this further, we profiled the application to obtain
hardware performance counter data. This revealed that the drop in
performance is due to a XXX

 With the
Intel compiler this drop in efficiency only occurs when going to
$512^2$. We therefore conclude that the drop in performance is due to
the Gnu compiler generating code that YYY
compared to that produced by the Intel and Cray
compilers.

For this compiler-friendly form of the code, there is generally little
performance difference between the executables produced by the Gnu and
Intel compilers on both the Haswell and Ivy Bridge CPUs (with the
exception of the $512^2$ case noted above). However, the executable
produced by the Cray compiler generally performs significantly better
for all but the largest domain size. Analysis of the optimisation
reports produced by the Cray and Intel compilers did not give any
clues as to the cause of this difference and an in-depth investigation
is outside the scope of this paper.

\subsection{Optimisations of the \psykal\ Version to Recover Performance}

We now examine the performance of the \psykal\ version of Shallow and
the code transformations that are required to reach the levels obtained
by the original.  We do this for the $256^{2}$ problem size since this
fits within last-level cache on all of the CPUs we are using here.
The performance of the \psykal\ version after each transformation has
been applied to the code is plotted in
Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
given relative to the performance of the original version of Shallow
with the same compiler/CPU.

\begin{figure}[!t]
\centering
\includegraphics[width=123mm]{opt_stages_256}
\caption{The performance of the \psykal\ version of Shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal\ version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called (module in-lining) has the
most dramatic effect on the performance of the compiled code; it now
achieves more than 80\% of the performance of the original. The
transformations required to get beyond this level then become
architecture dependent: on Haswell loop fusion achieves more than 90\%
while on Ivy Bridge it is fully-fusing the first loop nest. The
dramatic performance improvement obtained from the module-inlining
transformation demonstrates that the Gnu compiler is unable to
optimise over separate files.

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal\ version of
Shallow achieves some 80\% of the performance of the original
version. Only two code transformations were required to increase this
to $\sim$100\%. The first of these was the fusion of the computational
loops. However, we found that performance suffered (bar 5) when this
was extended to include the loop required for field copies at the end
of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
investigation of this slow-down revealed that the compiler was unable
to SIMD vectorise the newly-fused loop after it had in-lined the body
of the kernel called from within it. Once this in-lining was done by
hand (bar 6), the compiler was able to vectorise the loop and the
performance of the original version of Shallow was recovered on the
Haswell system. On Ivy Bridge the first loop nest also had to be fully
fused (bar 7) to recover $\sim100\%$ of the performance.

@@

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal\ version of Shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal\ version is fairly
good at 74\% of that of the original. In contrast to the other
compilers consisdered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal\ version that achieves 88\% of the
performance of the original.

Subsequent transformations actually harm performance until the
field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
with the preceeding loop. The resulting \psykal\ version now achieves
94\% of the original performance. Two further steps are required to
match the latter. First, the first loop nest has to be fully fused
(both inner and outer loops fused) and second, the array/loop bounds
are passed as an argument to the middle layer. This is done in order
to indicate to the compiler that these bounds remain constant for the
duration of each trip of the time-stepping loop.

Moving to the IBM compiler on the Power 8 system (sixth cluster in
Figure~\ref{FIG_opt_stages_256}) we see that performance of the
vanilla \psykal\ version is only 5\% of the original and several
transformations are required to achieve comparable performance. As
with the Cray compiler, specifying array bounds explicitly improves
performance. Module-inlining also has a significant effect; doubling
the performance from 10 to 20\% of the original. However, fusing loop
nests actually harms performance, just as it did for the Cray
compiler. Kernel inlining takes performance to over 50\% and it is the
IBM-supplied transformations of loop unrolling and fissioning the
first loop ({\it i.e.} undoing some of transformations three and
seven) that finally recovers 96\% of the performance of the
original. We note that this final pair of transformations significantly
reduce the performance obtained with the Intel and Gnu compilers
although they do not much perturb that obtained by the Cray
compiler.

\subsection{Performance Comparison of the Original and \psykal\ Versions}

Finally, we compare the performance of the best \psykal\ version of
Shallow with that of the original. In
Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
  difference} between the performance of the original and the (best)
\psykal\ versions of Shallow for each compiler/CPU combination. The
most important feature of this plot is that the performance penalty
incurred by the \psykal\ version is less than four percent with only
one exception where it is 6.6\%.  In fact, for the Intel and Cray
compilers, the \psykal\ version of Shallow is never more than 2\%
slower than the original and in some cases is faster.  This
demonstrates that we can reliably recover the performance of the
original version of the code, despite the significant restructuring
required by the \psykal\ approach.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=120mm]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal
version with that of the original version of the code. A negative value 
indicates that the \psykal\ version is faster than the original.}
\label{FIG_slowdown_summary}
\end{figure}

Turning to the cases where the \psykal\ version is significantly
{\em faster} than the original, we see from
Figure~\ref{FIG_slowdown_summary} that, with the exception of Power 8,
this only occurs for the smallest domain size. This is no doubt due in
part to the fact that the problem executes very quickly and therefore
a small reduction in run-time will show as a large precentage. Since
this problem is well contained in L2 cache, its performance will also
be more sensitive to small differences, {\it e.g.} in the way that CPU
registers are used.

The performance of the Power 8 system with the $512^2$ domain size is
somewhat anomolous: the \psykal\ version of the code performs
significantly better than the original and in addition, this problem
size is also the only one where the Power 8 out-performs the Haswell
system (see Figure~\ref{FIG_orig_perf_summary}) for the original
version.  The results in Figure~\ref{FIG_orig_perf_summary} indicate
that the $512^2$ case marks a transition from the working set being in
cache (sizes $\leq 256^2$) to spilling out of cache (size $1024^2$). We
therefore suggest that cache is being used more efficiently on the
Power 8 system and this is further enhanced by the IBM-supplied
transformations in the \psykal\ version.

@@

We have investigated the application of the \psykal\ separation of
concerns approach to the domain of finite-difference ocean
models. This approach enables the computational science (performance)
related aspects of a computer model to be kept separate from the
natural (oceanographic) science aspects.

@@

In particular, the results on the IBM system demonstrate two key points.
%
First, even for a relatively simple
benchmark code, the code structure required to obtain good performance
between different architecture and/or compiler combinations may
differ. For hand optimised codes this implies the need to support
multiple code bases in order to achieve portable performance.
%
Second, the \psykal\ approach directly addresses the problem of
requiring different optimisations for different architectures and/or
compilers by limiting these differences to a separate performance
layer, thereby allowing the natural science code base to remain
unchanged.

In future work we will extend our performance portability analysis of
the \psykal\ approach to shared memory parallelisation optimisations on
different architectures using different parallelisation techniques (in
particular OpenMP on multi/many-core and OpenACC on GPU's).
%
We will then analyse the performance of a domain-specific compiler
that is currently under development. This compiler will generate
optimised PSy-layer code using a user-provided recipe of
transformations, thereby removing the need for optimisation experts to
manually write the PSy layer.
